---
title: "Utility Burdens in Knoxville Households, hot sauce"
author: "Forest Gregg"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
font_adjustment: +1
always_allow_html: yes 
urlcolor: blue
---

```{r setup, echo = FALSE, message = FALSE}
library(knitr)
library(tidyverse)
library(tidycensus)
library(srvyr)

options(scipen=999, digits=2)
knitr::opts_chunk$set(echo = FALSE, cache = FALSE, message = FALSE, warning = FALSE, max.print = -1)
```

```{r define_functions}
get_knoxville_pums <- function(survey) {
    vars = c(
        "ELEFP", "ELEP", "HINCP", "GASFP", "GASP",
        "WATFP", "WATP", "TEN", "RELSHIPP", "RACBLK",
        "HISP", "RACWHT", "FULFP")
    knoxville_pums <- tidycensus::get_pums(
        variables = vars,
        state = 'TN',
        puma = c('01603'),
        survey = survey,
        year = 2019,
        recode = TRUE,
        rep_weights = "housing"
    )

    # We want households to be our units, but get_pums returns
    # one row per person, so we need to filter the data to
    # just one row per household. We'll filter the data to
    # select the row for the head of the household
    knoxville_pums <- knoxville_pums %>%
        filter(SPORDER == 1 & RELSHIPP_label == 'Reference person')

    # Once we have one row per household, we can convert the
    # data to a survey object, which will help us handle
    # sample weights and standard errors
    knoxville_pums <- tidycensus::to_survey(knoxville_pums,
                                            type='housing')

    # We are going to simplify the analysis by just looking
    # Black and white folks
    knoxville_pums <- knoxville_pums %>%
        filter((RACBLK_label == 'Yes' | RACWHT_label == 'Yes') & 
               HISP_label == 'Not Spanish/Hispanic/Latino')

    # Recode a missing value indicator
    knoxville_pums <- knoxville_pums %>%
        mutate(
            HINCP = ifelse(HINCP == -60000, NA, HINCP))

    knoxville_pums <- knoxville_pums %>%
        mutate(
            race = ifelse(RACWHT_label == 'Yes', 'white', 'Black'),
            race = fct_relevel(race, "white"))

    # The general practice in the literature is to code
    # Occupied without payment of rent as a renter, when comparing
    # owners and renters
    knoxville_pums <- knoxville_pums %>%
        mutate(
            tenure = case_when(TEN_label == 'Owned with mortgage or loan (include home equity loans)' ~ 'Owner Occupied',
                               TEN_label == 'Owned free and clear' ~ 'Owner Occupied',
                               TEN_label == 'Rented' ~ 'Renter Occupied',
                               TEN_label == 'Occupied without payment of rent' ~ 'Renter Occupied'))

    knoxville_pums <- knoxville_pums %>% 
        mutate(
            kub_energy = (case_when(GASFP_label == 'Valid monthly gas cost in GASP' ~ GASP,
                                    GASFP_label == 'No charge or gas not used' ~ 0,
                                    GASFP_label == 'Included in electricity payment' ~ 0,
                                    TRUE ~ NA_real_) * 12 +
                          case_when(ELEFP_label == 'Valid monthly electricity cost in ELEP' ~ ELEP,
                                    ELEFP_label == 'No charge or electricity not used' ~ 0,
                                    TRUE ~ NA_real_) * 12),
            kub_water = case_when(WATFP_label == 'Valid annual water cost in WATP' ~ WATP,
                                  WATFP_label == 'No charge' ~ 0,
                                  TRUE ~ NA_real_),
            kub_utilities = kub_energy + kub_water,
            kub_burden = kub_utilities / HINCP)

    # exclude households that use home heating oil
    knoxville_pums <- knoxville_pums %>%
        filter(FULFP_label == 'No charge or fuel other than gas or electricity not used')

    return(knoxville_pums)
}
```

```{r acs1, include=FALSE}
knoxville_pums <- get_knoxville_pums('acs1')

knoxville_pums_filtered <- knoxville_pums %>% 
    filter(kub_burden >=0 & kub_burden <=1)

tenure <- knoxville_pums_filtered %>%
    srvyr::survey_count(
        race,
        tenure,
        kub_burden > 0.1)
```

This report provides estimates of the proportion of Knoxville households that
are utility burdened, broken down by race and housing
tenure. We use the 2019 1-year American Community Survey 
Public Use Microdata Sample. These are a [sample of the individual surveys
conducted in 2019 for US Census's American Community Survey](https://www.census.gov/programs-surveys/acs/microdata.html).

We limit our analysis to households that are located in the City of Knoxville and
households where water, electric, or gas bills are not included in the rent or condo fees, but paid
separately. Households that use wood, heating oil, or something else other than gas or electric for heating are excluded. To simplify the analysis we consider only households headed by a Non-Hispanic
white person or Non-Hispanic Black person.

A household is utility burdened if the household's total water, gas, and 
electric bills is greater than 10% of the annual household income. The survey
asks about last month's costs for electricity and gas, and for past 12 months' cost
for water. Household income is reported annually. For this analysis, we annualize 
the monthly electric and gas bill costs. We only consider households where the utility burden is between 0% and 100% of their annual income. This excludes households
with negative income and some households that do not report credible utility bills or annual income.

```{r overall_table, results='asis'}    
knoxville_pums_filtered %>% 
    srvyr::survey_count(kub_burden > 0.1) %>%
    select("kub_burden > 0.1", n) %>%
    pivot_wider(
        names_from="kub_burden > 0.1", 
        values_from=n) %>%
    as.matrix() %>%
    proportions(margin=1) %>% 
    kable(col.names = c('Not utility burdened', 'Utility Burdened'),
          caption = 'Utility Burdened Households')
```


```{r black_non_black_table, results='asis'}    
knoxville_pums_filtered %>% 
    srvyr::survey_count(race, kub_burden > 0.1) %>%
    pivot_wider(
        race,
        names_from="kub_burden > 0.1", 
        values_from=n) %>%
    column_to_rownames('race') %>%
    as.matrix() %>%
    proportions(margin=1) %>% 
    kable(col.names = c('Not utility burdened', 'Utility Burdened'),
          row.names = TRUE,
          caption = 'Utility Burdened Households by Race')
```

```{r tenure_table, results='asis'}    
knoxville_pums_filtered %>% 
    srvyr::survey_count(tenure, kub_burden > 0.1) %>%
    pivot_wider(
        tenure,
        names_from="kub_burden > 0.1", 
        values_from=n) %>%
    column_to_rownames('tenure') %>%
    as.matrix() %>%
    proportions(margin=1) %>% 
    kable(col.names = c('Not utility burdened', 'Utility Burdened'),
          row.names = TRUE,
          caption = 'Utility Burdened Households by Tenure')
```


```{r black_non_black_owner, results='asis'}
tenure %>% 
    filter(tenure == 'Owner Occupied') %>%
    pivot_wider(
        race,
        names_from="kub_burden > 0.1",
        values_from=n) %>%
    column_to_rownames('race') %>%
    as.matrix() %>%
    proportions(margin=1) %>% 
    kable(col.names = c('Not utility burdened', 'Utility Burdened'),
          row.names = TRUE,
          caption = 'Utility Burdened Households by Race, Owner Occupied')
```

```{r black_non_black_renter, results='asis'}
tenure %>% 
    filter(tenure == 'Renter Occupied') %>%
    pivot_wider(
        race,
        names_from="kub_burden > 0.1",
        values_from=n) %>%
    column_to_rownames('race') %>%
    as.matrix() %>%
    proportions(margin=1) %>% 
    kable(col.names = c('Not utility burdened', 'Utility Burdened'),
          row.names = TRUE,
          caption = 'Utility Burdened Households by Race, Renter Occupied')
```

The differences in the proportion of households that are utility burdened between 
owner occupied households and renter occupied households and the differences between
Black and white households are all statistically significant at the 95% confidence level.

```{r model, include=FALSE}
model <- survey::svyglm(
    I(kub_burden > 0.1) ~ race + tenure, 
    design=knoxville_pums_filtered)
summary(model)
```
